---
title: "Predicting proficiency in weight lifting"
author: "Tomás E. Tecce"
output:
  html_document:
    theme: flatly
    highlights: pygments
---

# Source data

The data set for this project comes from
http://groupware.les.inf.puc-rio.br/har.  It consists of data from
accelerometers on the belt, forearm, arm, and dumbell of 6 participants,
who were asked to perform barbell lifts correctly and incorrectly in 5
different ways. First, I downloaded the training data set and load it into
R as a data frame.

```{r}
if (!file.exists("data")) { dir.create("data") }
if (!file.exists("data/pml-training.csv")) {
  fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  download.file(fileUrl, destfile="data/pml-training.csv", method="curl")
}

dftrain <- read.table("data/pml-training.csv", sep=",",
                      stringsAsFactors=FALSE, header=TRUE,
                      na.strings=c("","NA"))
```

# Feature selection

The training data set consists of `r dim(dftrain)[1]` observations of
`r dim(dftrain)[2]` variables. I start by checking for missing data.

```{r, echo=FALSE}
na_check <- as.data.frame(sapply(dftrain, function(x) { sum(is.na(x)) }))
```

Counting the number of NA values in each column I find that the data has
either columns with no missing data, or mostly empty columns (all with the
same value of `r max(100*na_check/dim(dftrain)[1])` per cent of NAs). I 
discard the latter columns from the training set, and also columns 1
through 7 which only contain data labels and time stamps.

```{r}
library(plyr)
library(dplyr)

features.list <- rownames(na_check)[na_check==0][-(1:7)]
dtrain <- select(dtrain, features.list)
```

